Asymmetric tail-association in extreme climatic events affects population-extinction risk

### 2019-10-11 MainText.Rmd
started editing intro...continued

---
title: "Asymmetric tail-dependence in extreme climatic events affects population-extinction risk"
author: "Shyamolina Ghosh, Daniel Reuman"
date: "`r format(Sys.time(), '%d %B, %Y')`"
geometry: "left=1in,right=1in,top=1in,bottom=1in"
output: 
  pdf_document:
    number_sections: no
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: head_maintext.sty

@@

```{r setup_MainText, echo=F}
library(rmarkdown)
knitr::opts_chunk$set(echo = TRUE, fig.pos = "H")
options(scipen = 1, digits = 5) #This option round all numbers appeared in the inline r code upto 5th digit
seed<-101
```

@@

@@

With global climate change, extreme climatic events (ECEs) become more frequent. These ECEs 
could pose severe impact on the resistance or resilience of a population and/ or community level through
life-history traits of the species. Therefore predicting the impact of extreme events 
on population extinction risk is necessery. We here introduce the potential use of "tail-dependence" 
property of two correlated ecological variables in modeling the extinction risk of 
an age-structured single patch model. We also extend this idea 
to density-dependent metapopulation models. For metapopulation models, 
we show extinction risk not only depends on the nature of tail-dependence of 
spatially correlated noise but is also influenced by the inherent dynamics 
of the population-model. To illaborate, we demonstrate left tail dependent noise across spatially correlated patches 
cause higher extinction risk than the right tail dependent noise, when the model show undercompensatory dynamics.
But in overcompensatory regime, the trend is opposite. Our study paves the path for future research considering this tail-dependence instead of simple correlation in population viability analysis.

\textbf{\textit{Keywords:}} extinction risk, extreme events, tail dependence, density-dependent model

@@

@@

<!--how common are the extreme events?-->
For past few decades, extreme climatic events (ECEs) appeared more common than 
usual due to anthropogenic climate change, e.g. temperature or precipitation extremes,
heat-waves, wild-fire etc. [@ummenhofer2017extreme]. Frequent extreme events 
can shift temperature or precipitation-mean, 
change variability of the distribution, as well as change its 
symmetry or skewness [@hansen2012perception]. Extreme climatic events can pose severe
impacts at the individual, population, community and ecosystem level [@bragazza2008climatic; 
@gutschick2003extreme; @jiguet2011community; @mckechnie2009climate;
@hoover2014resistance]. Therefore, 
understanding the long-term impacts of changes 
in ECEs has been a longstanding interest in ecological research 
[@katz2005statistics; @smith2011ecological; @diez2012will; @moritz2013future; @bailey2016tackling].

<!--How extinction risk is related to extreme events : past studies
consequences of extreme events on population life history, community-->
Earlier studies showed ECEs may affect species abundances (species re-ordering) 
<!--***DAN: Shyamolina, I do not know what the parenthesis refers to -->
<!--***DAN: Shyamolina, you have obviously done a lot of reading to situate your manuscript
in the appropriate literature, and that is great!-->
through 
life-history traits, alter community structures through cascading effect on 
higher hierarchical levels and lead to local extinction of 
species [@thomas2004extinction; @edgar2010nino; @moreno2011extreme]. 
<!--Should I mention idea of match-mismatch hypothesis?-->
For example, a study [@chiu2013impact] showed extreme flood affected 
the annual survival of a prey community that in turn affected predators in 
aquatic as well as adjacent riparian habitats via trophic linkages. Another study by 
@frederiksen2008demographic used $43$ years of ring-recovery data 
to examine the demographic impact of ECEs 
through temporal variability in juvenile, immature and adult survival
of European shags on the Isle of May in eastern Scotland.
A relatively recent study [@mcdermott2017sensitivity] investigated the sensitivity 
of $41$ UK butterfly species to four different ECEs (extreme heat, extreme cold, 
extreme precipitation, drought) over $37$ years at the site level, across 
each species' life-stages. Experimental study showed successive ECEs with short time intervals 
could weaken the resistance of plant communities too [@dreesen2014successive]. 
All these studies indicated that environmental stochasticity (through ECEs) 
can lead to demographic stochasticity (through influence on reproductive or 
mortality rate) of different age class of a population making it vulnerable 
to local extinction. 
<!--***DAN: Shyamolina, demogra[hic stochasticity is a specific thing. I suspect you may
just want to say instead something like "All these studies indicated that environmental stochasticity (through ECEs) can make a population vulnerable to extirpation."] -->

<!--Earlier study on metapopulation extinction risk : Large-scale spatial synchrony may also be generated by
other processes such as migration and predation, but regional
stochasticity in the form of spatially correlated weather conditions is
probably the dominant synchronizing mechanism (ref: Nature1998) but what about taildep. effect on extrisk?.-->
In the metapopulation context, regional stochasticity 
can simultaneously affected multiple habitat patches where spatially correlated weather conditions were 
probably the dominant synchronizing mechanism for extinctions [@hanski1998metapopulation]. A 21 years-long
study by @tack2015increasing on the Glanville fritillary butterfly demonstrated that frequency 
of drought during early larval instars increased amplitude and span of spatial synchrony 
over a metapopulation network of 4000 dry meadows and could affect the long-term population viability.
Though earlier literature considered the effect of spatial and/ or temporal correlation in estimating extinction eisk of a metapopulation, but to our knowledge, using the concept of "tail dependence" (i.e. correlation would be stronger during extreme years) to model extinction risk remains unexplored.
<!--***DAN: Shyamolina, you cannot intro tail dependence (please change to calling it
tail association) in the last sentence of a paragraph. The reader does not know what that is, 
and a whole paragraph is needed to intro it.-->

<!--***DAN: Shyamolina, you need to make two major changes in how tail associations
are introdcued. 
  First, do it prior to saying anything about copulas. We have been wanting
to disassociate this paper, to some extent, from BIVAN. Well, probably the best way
to do that is by introducing tail dependence first, before copulas. You can 
describe it intuitively as association in the tails, and you can use pictures to describe it.
You can introduce the partial Spearman correlation (briefly in the Intro, more detail in Methods). 
You will need an explnation of why we 
use Spearman instead of Pearson (talk to me if you do not know, but you may want to read the 
new Background section of BIVAN again first). One does not need to know anything about copulas
to understand tail associations.
  Second, the introduction to copulas that is given is outmoded. See the BAckground section
to BIVAN for improvements. But you do not want to say anything like that much about copulas 
here, you just want to briefly link to the concept, cite some of the Intro work that
has been done (includeing BIVAN and some of the paper BIVAN cites for Intro work),
mention that tail associations are one aspect of copula structure that we have found important,
and introduce any specific things you will need, like the Clayton copula.

@@

<!--How copula could be relevent in estimating extreme event's 
influence on extinction risk : introduce Taildep. concept-->
We recently [@Ghosh_copula] used this tail dependence feature using "copula" - a statistical 
tool broadly used in modeling dependence between variables - to estimate extinction risk of a density-independent 
metapopulation model. Present manuscript is an extension of this concept 
in estimating the extinction risk beyond the density-independent model 
and even applicable for a simple age-structured population where its vital rates are influenced by ECEs. 
Following we will reintroduce the concept of copula in a manner that follows @Ghosh_copula.

@@

<!--tail dep. of copula & extinction risk-->
A major benefit of the copula approach is that it can effectively capture the tail dependence. 
Tail dependence, already exemplified in Fig..... (add Fig.), is association between the more extreme 
values of distributions. For two positively associated variables, left-tail 
(respectively, right-tail) dependence is stronger association between
values in the left or lower portions, i. e. for smaller values 
(respectively, the right or upper portions, i.e. for higher values) 
of the two distributions. Stronger dependence in either tail causes asymmetric copula 
structure for the joint distribution of two variables. For example, Clayton and Survival Clayton are 
two copula families 
<!--***DAN: Shyamolina, we cannot just mention these, no one will know what they are.
As I said I think we should intro tail associations without copulas, and then subsequently link 
to copulas afterwards.-->
with asymmetric tail that are used to model exclusively 
left and right tail dependence between variables, respectively.
From ecological perspective, it is reasonable to expect that extreme cold winter during growing season
might reduce the adult reproduction rate as well as larval survival (i.e. stronger 
tail dependence between vital rates when temperature is too cold than normal) and 
thereby inflate the population viability. Similarly, in context of 
density-dependent metapopulation models, if extreme environmental fluctuations across sites 
appear to be as synchronous (left tail dependent or right tail dependent between patches 
with same correlation as in Fig.....) then it could wipe out some metapopulation patches 
during unfavorable years unless they are rescued by the nearby populations. This scenario 
could be more complex and specific of the direction of dependence 
depending on the dynamics of the model (see \nameref{Results}). Though tail-dependence 
property of a copula has been used for risk-analysis in other fields like finance, 
hydrology etc. [@alexander2008market; @li2013; @chen2011flood; @Goswami2018], but to our 
knowledge asymmetric tail dependence have never been considered 
in estimating the effect of ECEs through extinction risk of age structured 
model or density-dependent metapopulations.

<!--our objective ?-->
In this paper, we will explore the potential use of asymmetric tail-dependence 
for understanding the extinction risk of populations towards extreme climatic event
by addressing the following questions: (Q1) Does nature of tail dependence 
(left tail or right tail) in vital rates
of a single-patch age-structured population model affect extinction risk of that population differently?
We hypothesized population might have higher extinction risk if vital rates are left-tail dependent.
(Q2) How does environmental noises with asymmetric tail dependence across multiple patches
influence extinction risk of a density-dependent metapopulation model? Our speculation 
is that the general finding from density-independent metapopulation model [@Ghosh_copula]
(higher extinction risk if noises across patches were left-tail dependent than the one 
if noises had right-tail dependence) would hold true only in the under-compensatory regime, but 
should be reverse if the model shows over-compensatory dynamics. 
(Q3) We further hypothesize that when there is no dispersal, the above-said switching should occur
near the critical parameter value where the deterministic model shows transition from 
under-compensatory to over-compensatory dynamics.

@@

<!--stochastic matrix model : age structured-->
To investigate (Q1) from \nameref{Introduction} that how does asymmetric tail 
dependent noise affect the extinction risk of
populations in a single patch age-structured model,
we considered a stochastic age-structured model (Eq. \ref{eq_age_str}) 
with two age classes (egg $E$ and female adult $A$). We assumed
the population was semelperous, i.e., adult ($A$) went through the
reproductive episode only once before death. Eggs and adults in the 
next generation evolved depending on the fecundity of adult ($f_a \geq 0$) 
and survival of eggs ($0 \leq s_e \leq 1$) from present generation.
Fluctuating vital rates ($f_a$ and $s_e$) captured the environmental
stochasticity (Eq. \ref{eq_vitalrates}). We made adult's fecundity 
($f_a$) negative density-dependent with its own population. When $\Omega = \infty$ in 
Eq. \ref{eq_vitalrates}, the projection matrix was independent of populations 
(i.e., Eq. \ref{eq_age_str} was "linear" model then), 
but when $\Omega$ was finite, Eq. \ref{eq_age_str} became "nonlinear".
Fecundity ($f_a$) and survival ($s_e$) rates were random draw from a gamma ($\Gamma(k,\theta)$) and
a beta ($\mathcal{B}(\alpha,\beta)$) distribution, respectively. We generated those distributions 
from iid noises with a varied degree of asymmetric tail-dependence between 
fecundity ($f_a$) and survival ($s_e$) rates. Degree of tail dependence had two broad 
categories: moderate and extreme. Each category had specific direction in tail dependence: 
either in left or in right tail of vital rates. Vital rates ($f_a$ and $s_e$) were
left tail dependent when unusual bad years ("catastrophs") suppressed both vital rates synchronously, 
and right tail dependent when unexpected good years ("bonanzas") boosted both of them.

@@

We used bivariate Clayton copula to model moderate left-tail and Survival Clayton copula to model 
moderate right tail dependence in the vital rates. Both copula had the same Kendall 
correlation ($=0.6$) but opposite dependence in their respective tails. 
We generated a $N \times 2$ matrix, M, for each of the left (Clayton) and 
right (Survival Clayton) tail-dependent bivariate copulas (see `BiCopSim` function from
`VineCopula` package ), where $N$ is number of years ($=25$) multiplied 
by number of simulations ($=10000$) we used. 
We chose parameters for each copula family using 
`iTau` function from `copula` package so that both copula had
same Kendall's correlation ($\tau = 0.6$). Entries within each clumn of
matrix $M$, then became independent of each others but 
entries between columns had a specific tail dependence
with positive Kendall's correlation ($\tau = 0.6$).

@@

<!--density-dependence metapopulation model-->
We explored (Q2) by examining how nature of the tail dependence of an
environmental variable across space could influence the extinction risk of a 
metapopulation model with density-dependence feedback. We considered six density-dependent 
population-growth models (Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, Austin-Brewer)
in noisy environment (see Table \ref{tab_model_summary}) 
from a metapopulation context.
Models were adapted from earlier work
by @cohen1995unexpected, excluding only two : Varley model (this is not a continuous one) 
and Malthus-Condorcet-Mill model (terms containing logarithm of extinct population does not make sense). 
All models assumed a generalized form : $\vec{P}_{t+1} = D \lambda_t \vec{P_t}$,
where $\vec{P_t}$ was a vector of population densities, 
the $i^{th}$ component $P_t^i$ representing
the density in the $i^{th}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{th}$ diagonal entry 
of stochastic density dependence form $\lambda_i(t)$ (see second column 
of Table \ref{tab_model_summary}). For simplicity, 
we considered $n$ patches in each model were arranged, evenly spaced, in a line
with the provision of "local" or "global" dispersal among patches. A n $\times$ n dispersal
matrix, $D$ assumed that a fraction $d$ of 
each population patch dispersed during any given time step, equally distributed to
the two or one nearest neighbors in case of local dispersal. For global dispersal, the fraction
$d$ dispersed equally to other $(n-1)$ patches. We called $d$ as dispersal
rate with $d$ being zero there was no dispersal, and hence no chance to rescue an extinct 
patch through migration from nearby metapopulation patches.

We simulated our metapopulation models with noises $(\epsilon_t^i)$ that were 
independent through time with same spatial correlations between patches (to within sampling error)
for every simulation, but moderately (or extremely) left- and right-tail dependent between patches in
different simulations. For moderately left-tail dependent noise we drew points from a $n$ dimensional 
Clayton copula with same Kendall correlation ($= 0.6$) between any two out of $n$ patches and made 
the marginals normal with zero mean and standard deviation $\sigma$ 
(see the specified value $\sigma$ for each model in 3$^{rd}$ 
column of Table \ref{tab_model_summary}). For moderately 
right-tail dependent noise we flipped ($180^{\circ}$ rotation) the points drawn 
from Clayton copula so that all the statistical properties remained 
same except for the direction of tail-dependence. We generated
extreme left- or right-tail dependent noises via algorithm stated in 
Appendix \ref{SM-extr_tail_dep}. Our models (Table \ref{tab_model_summary}) 
computed extinction risk after $25$ years for $10000$ simulations at $n = 5$ patches
with spatially correlated noises having either moderate or extreme tail 
dependence in left and right-tail. <!--say how you set ext. threshold = Pe/10-->
We started with initial population $P_{t=0}^i = P_e^i, (i = 1,2,\ldots,n)$ for each patch, where 
$P_e^i$ was the equilibrium value (or similar to carrying capacity concept) 
for each density-dependence model (see 4$^{th}$ column of Table \ref{tab_model_summary}).
We assumed population went extinct after each time step if it was below a 
certain threshold (set to 10\% of initial population $P_e$ for individual model).

@@

For single patch linear age-structured models, we found higher extinction risk for moderate left-tail 
dependence (modeled by Clayton copula) in fluctuating vital 
rates compared with the one when vital rates were right tail dependent (modeled by Survival Clayton copula).
It was expected because an extreme "bad" year (e.g. with severe cold) can suppress 
the fecundity of adult and survival/ maturation of eggs simultaneously 
and thereby can increase the extinction risk of the population. For non-linear age-structured 
models, we got the similar trend for extinction risks with different tail-dependence (left and right).
Non-linear model showed higher extinction risk than its corresponding 
linear analog as it was evident from Eq. \ref{eq_vitalrates} with finite $\Omega = 100$. 
Fig. \ref{SM-fig_extrisk_agestr_mtd} represents this scenario with two copulas having different 
asymmetric tails but same Kendall's correlation ($\tau = 0.6$). Extinction risks for
two different tail-dependent vital rates were not strinkingly different (e.g. see pair of red and
blue lines in Fig. \ref{SM-fig_extrisk_agestr_mtd}A or B) and we anticipated the difference being partly 
dependent on the degree of tail dependence. Therefore we repeated the same analysis with 
vital rates coming from extreme-tail dependent (left and right in two different simulations)
copulas and got the similar results but with much prominent difference between their extinction risks
(Fig. \ref{fig_extrisk_agestr_etd}). This result answered (Q1) and showed extinction risk of 
a population could be considered as direct manifestation of tail-dependence property 
between vital rates of different age-class
of that population. <!--discussion later: add Tuljapurkar's approximation-->

@@

tail-dependence effect on extinction risk 

>>>>

---
title: "Asymmetric tail-association in extreme climatic events affects population-extinction risk"
author: "Shyamolina Ghosh, Daniel Reuman"
date: "`r format(Sys.time(), '%d %B, %Y')`"
geometry: "left=1in,right=1in,top=1in,bottom=1in"
output: 
  pdf_document:
    number_sections: no
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: head_maintext.sty

@@

```{r setup_MainText, echo=F}
library(rmarkdown)
knitr::opts_chunk$set(echo = TRUE, fig.pos = "H")
options(scipen = 1, digits = 3) #This option round all numbers appeared in the inline r code upto 5th digit
seed<-101
```

@@

# Alternative titles
1) Tail associations in environmental variables and extinction risk
2) The influence of correlations in the extremes on extinction risk

@@

With global climate change, extreme climatic events (ECEs) become more frequent. These ECEs 
could pose severe impact on the resistance or resilience of a population and/ or community level through
life-history traits of the species. Therefore predicting the impact of extreme events 
on population extinction risk is necessery. We here introduce the potential use of "tail-association" 
property of two correlated ecological variables in modeling the extinction risk of 
an age-structured single patch model. We also extend this idea 
to density-dependent metapopulation models. For metapopulation models, 
we show extinction risk not only depends on the nature of tail-association of 
spatially correlated noise but is also influenced by the inherent dynamics 
of the population-model. To illaborate, we demonstrate left tail dependent noise across spatially correlated patches 
cause higher extinction risk than the right tail dependent noise, when the model show undercompensatory dynamics.
But in overcompensatory regime, the trend is opposite. Our study paves the path for future research considering this tail-association instead of simple correlation in population viability analysis.

\textbf{\textit{Keywords:}} extinction risk, extreme events, tail association, density-dependent model

@@

```{r read_res, echo=F}
ktau_common<-readRDS("./Results/ktau_common.RDS")
```

@@

<!--how common are the extreme events?-->
For past few decades, extreme climatic events (ECEs) appeared more common than 
usual due to anthropogenic climate change, e.g. temperature or precipitation extremes,
heat-waves, wild-fire etc. [@ummenhofer2017extreme]. Frequent extreme events 
can shift temperature or precipitation-mean, 
change variability of the distribution, as well as change its 
symmetry or skewness [@hansen2012perception]. Extreme climatic events can pose severe
impacts at the individual, population, community and ecosystem level [@bragazza2008climatic; 
@gutschick2003extreme; @jiguet2011community; @mckechnie2009climate;
@felton2017integrating]. Therefore, 
understanding the long-term impacts of changes 
in ECEs has been a longstanding interest in ecological research 
[@katz2005statistics; @smith2011ecological; @diez2012will; @moritz2013future; @bailey2016tackling].

<!--How extinction risk is related to extreme events : past studies
consequences of extreme events on population life history, community-->
Earlier studies showed ECEs may affect species abundances (species re-ordering) 
<!--***DAN: Shyamolina, I do not know what the parenthesis refers to

***Shyamolina: species reordering is used to access the change in community 
in response to extreme events or global environmental changes, it relates to change 
in relative abundances of dominant species (mainly). For example, as different species have different 
sensitivity levels to extreme condition (means the extreme threshold is not 
constant for all the species in a community), some years with extreme drought favors say c3 grass 
but in the next year c4 grass could be dominant. Such alteration in community composition has "impact on 
ecosystem function, leading to an extreme ecological response". For detailed overview you may see 
@hoover2014resistance, Jones et al. Ecol. lett. 2017 (doi: 10.1111/ele.12864), 
Smith et al., Ecology 2009 (doi:10.1890/08-1815.1).
Actually LAN K. KNAPP and MELINDA D. SMITH did a lot of work on this to access changes in 
plant community response to extreme events 
(this review article: Felton et al. 2017, Phil. Trans. R. Soc. B 372: 20160142 is a nice summary).
-->
<!--***DAN: Shyamolina, you have obviously done a lot of reading to situate your manuscript
in the appropriate literature, and that is great!-->
through 
life-history traits, alter community structures through cascading effect on 
higher hierarchical levels and lead to local extinction of 
species [@thomas2004extinction; @edgar2010nino; @moreno2011extreme; @hoover2014resistance]. 
<!--Should I mention idea of match-mismatch hypothesis?-->
For example, a study [@chiu2013impact] showed extreme flood affected 
the annual survival of a prey community that in turn affected predators in 
aquatic as well as adjacent riparian habitats via trophic linkages. Another study by 
@frederiksen2008demographic used $43$ years of ring-recovery data 
to examine the demographic impact of ECEs 
through temporal variability in juvenile, immature and adult survival
of European shags on the Isle of May in eastern Scotland.
A relatively recent study [@mcdermott2017sensitivity] investigated the sensitivity 
of $41$ UK butterfly species to four different ECEs (extreme heat, extreme cold, 
extreme precipitation, drought) over $37$ years at the site level, across 
each species' life-stages. Experimental study showed successive ECEs with short time intervals 
could weaken the resistance of plant communities too [@dreesen2014successive]. 
All these studies indicated that environmental stochasticity (through ECEs) 
can lead to demographic stochasticity (through influence on reproductive or 
mortality rate) of different age class of a population making it vulnerable 
to local extinction. 
<!--***DAN: Shyamolina, demogra[hic stochasticity is a specific thing. I suspect you may
just want to say instead something like "All these studies indicated that environmental stochasticity (through ECEs) can make a population vulnerable to extirpation."] -->

<!--Earlier study on metapopulation extinction risk : Large-scale spatial synchrony may also be generated by
other processes such as migration and predation, but regional
stochasticity in the form of spatially correlated weather conditions is
probably the dominant synchronizing mechanism (ref: Nature1998) but what about taildep. effect on extrisk?.-->
In the metapopulation context, regional stochasticity 
can simultaneously affected multiple habitat patches where spatially correlated weather conditions were 
probably the dominant synchronizing mechanism for extinctions [@hanski1998metapopulation]. A 21 years-long
study by @tack2015increasing on the Glanville fritillary butterfly demonstrated that frequency 
of drought during early larval instars increased amplitude and span of spatial synchrony 
over a metapopulation network of 4000 dry meadows and could affect the long-term population viability.
Therefore, understanding the synchronized fluctuation over a metapopulation-network
remains important to the conservationists because well connected patches, if synchronous, are more 
prone to simultaneous extinction during extreme bad years [@earn2000coherence].

<!--concept for correlation in the extreme tail: tail-association-->
Though earlier literature considered the influence of spatial and/ or temporal 
correlations in environmental variables on extinction risk of a metapopulation, to 
our knowledge the separate influence on extinction risk of 'tail associations' between
environmental time series in different locations has hardly been considered.

” But this is not the topic sentence. The topic sentence comes second and is something like “Tail associations are…” or “We say two positively associated variables are more right- than left-tail associated if…”. Then the para continues, its point being to explain what asymmetric tail associations are. You can use a pedagogical figure. One thing you probably want to include is the logic (see BIVAN) that if extreme events are getting stronger and more extreme/intense events are also more spatially widespread, then it seems reasonable to expect that spatial patterns of tail associations in environmental variables will also be changing. You may also want to mention (one sentence only) that we use the statistical concept of the copula to control tail dependence in our simulations – more will be said about copulas in the Methods.

<!--***DAN: Shyamolina, you need to make two major changes in how tail associations
are introdcued. 
  First, do it prior to saying anything about copulas. We have been wanting
to disassociate this paper, to some extent, from BIVAN. Well, probably the best way
to do that is by introducing tail dependence first, before copulas. You can 
describe it intuitively as association in the tails, and you can use pictures to describe it.
You can introduce the partial Spearman correlation (briefly in the Intro, more detail in Methods). 
You will need an explnation of why we 
use Spearman instead of Pearson (talk to me if you do not know, but you may want to read the 
new Background section of BIVAN again first). One does not need to know anything about copulas
to understand tail associations.

  Second, the introduction to copulas that is given is outmoded. See the BAckground section
to BIVAN for improvements. But you do not want to say anything like that much about copulas 
here, you just want to briefly link to the concept, cite some of the Intro work that
has been done (includeing BIVAN and some of the paper BIVAN cites for Intro work),
mention that tail associations are one aspect of copula structure that we have found important,
and introduce any specific things you will need, like the Clayton copula.

@@

<!--How copula could be relevent in estimating extreme event's 
influence on extinction risk : introduce Taildep. concept-->
We recently [@Ghosh_copula] used this tail association feature using "copula" - a statistical 
tool broadly used in modeling dependence between variables - to estimate extinction risk of a density-independent 
metapopulation model. Present manuscript is an extension of this concept 
in estimating the extinction risk beyond the density-independent model 
and even applicable for a simple age-structured population where its vital rates are influenced by ECEs. 
Following we will reintroduce the concept of copula in a manner that follows @Ghosh_copula.

@@

<!--tail dep. of copula & extinction risk-->
A major benefit of the copula approach is that it can effectively capture the tail association. 
Tail association, already exemplified in Fig..... (add Fig.), is association between the more extreme 
values of distributions. For two positively associated variables, left-tail 
(respectively, right-tail) association is stronger association between
values in the left or lower portions, i. e. for smaller values 
(respectively, the right or upper portions, i.e. for higher values) 
of the two distributions. Stronger association in either tail causes asymmetric copula 
structure for the joint distribution of two variables. For example, Clayton and Survival Clayton are 
two copula families 
<!--***DAN: Shyamolina, we cannot just mention these, no one will know what they are.
As I said I think we should intro tail associations without copulas, and then subsequently link 
to copulas afterwards.-->
with asymmetric tail that are used to model exclusively 
left and right tail association between variables, respectively.
From ecological perspective, it is reasonable to expect that extreme cold winter during growing season
might reduce the adult reproduction rate as well as larval survival (i.e. stronger 
tail association between vital rates when temperature is too cold than normal) and 
thereby inflate the population viability. Similarly, in context of 
density-dependent metapopulation models, if extreme environmental fluctuations across sites 
appear to be as synchronous (left tail dependent or right tail dependent between patches 
with same correlation as in Fig.....) then it could wipe out some metapopulation patches 
during unfavorable years unless they are rescued by the nearby populations. This scenario 
could be more complex and specific of the direction of association 
depending on the dynamics of the model (see \nameref{Results}). Though tail-association 
property of a copula has been used for risk-analysis in other fields like finance, 
hydrology etc. [@alexander2008market; @li2013; @chen2011flood; @Goswami2018], but to our 
knowledge asymmetric tail association have never been considered 
in estimating the effect of ECEs through extinction risk of age structured 
model or density-dependent metapopulations.

<!--our objective ?-->
In this paper, we will explore the potential use of asymmetric tail-association 
for understanding the extinction risk of populations towards extreme climatic event
by addressing the following questions: (Q1) Does nature of tail association 
(left tail or right tail) in vital rates
of a single-patch age-structured population model affect extinction risk of that population differently?
We hypothesized population might have higher extinction risk if vital rates are left-tail dependent.
(Q2) How does environmental noises with asymmetric tail association across multiple patches
influence extinction risk of a density-dependent metapopulation model? Our speculation 
is that the general finding from density-independent metapopulation model [@Ghosh_copula]
(higher extinction risk if noises across patches were left-tail dependent than the one 
if noises had right-tail association) would hold true only in the under-compensatory regime, but 
should be reverse if the model shows over-compensatory dynamics. 
(Q3) We further hypothesize that when there is no dispersal, the above-said switching should occur
near the critical parameter value where the deterministic model shows transition from 
under-compensatory to over-compensatory dynamics.

<!--FIG : pedagogical fig showing 4 types of tail associations: 
extreme left, moderate left, moderate right, extreme right-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/introfig_copula.pdf}
\caption{Two random variables can have same Kendall 
correlation ($\tau=$ `r ktau_common`), but different degree of associations
in their extreme values: (A) extreme left tail, (B) moderate left tail, (C) moderate right tail, and
(D) extreme right tail association.\label{fig_intro_cop}}
\end{center}
\end{figure*}

@@

<!--stochastic matrix model : age structured-->
To investigate (Q1) from \nameref{Introduction} that how does asymmetric tail 
dependent noise affect the extinction risk of
populations in a single patch age-structured model,
we considered a stochastic age-structured model (Eq. \ref{eq_age_str}) 
with two age classes (egg $E$ and female adult $A$). We assumed
the population was semelperous, i.e., adult ($A$) went through the
reproductive episode only once before death. Eggs and adults in the 
next generation evolved depending on the fecundity of adult ($f_a \geq 0$) 
and survival of eggs ($0 \leq s_e \leq 1$) from present generation.
Fluctuating vital rates ($f_a$ and $s_e$) captured the environmental
stochasticity (Eq. \ref{eq_vitalrates}). We made adult's fecundity 
($f_a$) negative density-dependent with its own population. When $\Omega = \infty$ in 
Eq. \ref{eq_vitalrates}, the projection matrix was independent of populations 
(i.e., Eq. \ref{eq_age_str} was "linear" model then), 
but when $\Omega$ was finite, Eq. \ref{eq_age_str} became "nonlinear".
Fecundity ($f_a$) and survival ($s_e$) rates were random draw from a gamma ($\Gamma(k,\theta)$) and
a beta ($\mathcal{B}(\alpha,\beta)$) distribution, respectively. We generated those distributions 
from iid noises with a varied degree of asymmetric tail-association between 
fecundity ($f_a$) and survival ($s_e$) rates. Degree of tail association had two broad 
categories: moderate and extreme. Each category had specific direction in tail association: 
either in left or in right tail of vital rates. Vital rates ($f_a$ and $s_e$) were
left tail dependent when unusual bad years ("catastrophs") suppressed both vital rates synchronously, 
and right tail dependent when unexpected good years ("bonanzas") boosted both of them.

@@

We used bivariate Clayton copula to model moderate left-tail and Survival Clayton copula to model 
moderate right tail association in the vital rates. Both copula had the same Kendall 
correlation ($=0.6$) but opposite association in their respective tails. 
We generated a $N \times 2$ matrix, M, for each of the left (Clayton) and 
right (Survival Clayton) tail-dependent bivariate copulas (see `BiCopSim` function from
`VineCopula` package ), where $N$ is number of years ($=25$) multiplied 
by number of simulations ($=10000$) we used. 
We chose parameters for each copula family using 
`iTau` function from `copula` package so that both copula had
same Kendall's correlation ($\tau = 0.6$). Entries within each clumn of
matrix $M$, then became independent of each others but 
entries between columns had a specific tail association
with positive Kendall's correlation ($\tau = 0.6$).

@@

<!--density-dependence metapopulation model-->
We explored (Q2) by examining how nature of the tail association of an
environmental variable across space could influence the extinction risk of a 
metapopulation model with density-dependence feedback. We considered six density-dependent 
population-growth models (Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, Austin-Brewer)
in noisy environment (see Table \ref{tab_model_summary}) 
from a metapopulation context.
Models were adapted from earlier work
by @cohen1995unexpected, excluding only two : Varley model (this is not a continuous one) 
and Malthus-Condorcet-Mill model (terms containing logarithm of extinct population does not make sense). 
All models assumed a generalized form : $\vec{P}_{t+1} = D \lambda_t \vec{P_t}$,
where $\vec{P_t}$ was a vector of population densities, 
the $i^{th}$ component $P_t^i$ representing
the density in the $i^{th}$ habitat patch ($i=1,2,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{th}$ diagonal entry 
of stochastic density dependence form $\lambda_i(t)$ (see second column 
of Table \ref{tab_model_summary}). For simplicity, 
we considered $n$ patches in each model were arranged, evenly spaced, in a line
with the provision of "local" or "global" dispersal among patches. A n $\times$ n dispersal
matrix, $D$ assumed that a fraction $d$ of 
each population patch dispersed during any given time step, equally distributed to
the two or one nearest neighbors in case of local dispersal. For global dispersal, the fraction
$d$ dispersed equally to other $(n-1)$ patches. We called $d$ as dispersal
rate with $d$ being zero there was no dispersal, and hence no chance to rescue an extinct 
patch through migration from nearby metapopulation patches.

We simulated our metapopulation models with noises $(\epsilon_t^i)$ that were 
independent through time with same spatial correlations between patches (to within sampling error)
for every simulation, but moderately (or extremely) left- and right-tail dependent between patches in
different simulations. For moderately left-tail dependent noise we drew points from a $n$ dimensional 
Clayton copula with same Kendall correlation ($= 0.6$) between any two out of $n$ patches and made 
the marginals normal with zero mean and standard deviation $\sigma$ 
(see the specified value $\sigma$ for each model in 3$^{rd}$ 
column of Table \ref{tab_model_summary}). For moderately 
right-tail dependent noise we flipped ($180^{\circ}$ rotation) the points drawn 
from Clayton copula so that all the statistical properties remained 
same except for the direction of tail-association. We generated
extreme left- or right-tail dependent noises via algorithm stated in 
Appendix \ref{SM-extr_tail_dep}. Our models (Table \ref{tab_model_summary}) 
computed extinction risk after $25$ years for $10000$ simulations at $n = 5$ patches
with spatially correlated noises having either moderate or extreme tail 
association in left and right-tail. <!--say how you set ext. threshold = Pe/10-->
We started with initial population $P_{t=0}^i = P_e^i, (i = 1,2,\ldots,n)$ for each patch, where 
$P_e^i$ was the equilibrium value (or similar to carrying capacity concept) 
for each density-dependence model (see 4$^{th}$ column of Table \ref{tab_model_summary}).
We assumed population went extinct after each time step if it was below a 
certain threshold (set to 10\% of initial population $P_e$ for individual model).

@@

For single patch linear age-structured models, we found higher extinction risk for moderate left-tail 
association (modeled by Clayton copula) in fluctuating vital 
rates compared with the one when vital rates were right tail dependent (modeled by Survival Clayton copula).
It was expected because an extreme "bad" year (e.g. with severe cold) can suppress 
the fecundity of adult and survival/ maturation of eggs simultaneously 
and thereby can increase the extinction risk of the population. For non-linear age-structured 
models, we got the similar trend for extinction risks with different tail-association (left and right).
Non-linear model showed higher extinction risk than its corresponding 
linear analog as it was evident from Eq. \ref{eq_vitalrates} with finite $\Omega = 100$. 
Fig. \ref{SM-fig_extrisk_agestr_mtd} represents this scenario with two copulas having different 
asymmetric tails but same Kendall's correlation ($\tau = 0.6$). Extinction risks for
two different tail-dependent vital rates were not strinkingly different (e.g. see pair of red and
blue lines in Fig. \ref{SM-fig_extrisk_agestr_mtd}A or B) and we anticipated the difference being partly 
dependent on the degree of tail association. Therefore we repeated the same analysis with 
vital rates coming from extreme-tail dependent (left and right in two different simulations)
copulas and got the similar results but with much prominent difference between their extinction risks
(Fig. \ref{fig_extrisk_agestr_etd}). This result answered (Q1) and showed extinction risk of 
a population could be considered as direct manifestation of tail-association property 
between vital rates of different age-class
of that population. <!--discussion later: add Tuljapurkar's approximation-->

@@

tail-association effect on extinction risk 
